raptr: Representative and Adequate Prioritisation Toolkit in R

### 2016-07-18 code/rmarkdown/article.Rmd
+ assorted ms fixes including Rich's edits + results analysed + change to raptr

---
title: "rapr: Representative and Adequate Prioritisations in R"
author: |
  | Jeffrey O. Hanson$^1$, Jonathan R. Rhodes$^2$, Hugh P. Possingham$^1$, Richard A. Fuller$^1$
  | $^1$School of Biological Sciences, The University of Queensland, Brisbane, QLD, Australia
  | $^2$School of Geography, Planning and Environmental Management, The University of Queensland, Brisbane, QLD, Australia
  | Correspondance should be addressed to jeffrey.hanson@uqconnect.edu.au
date: "`r format(Sys.time(), '%d %B %Y')`"
tags: ["conservation", "biodiversity", "MILP", "reserve-selection", "optimisation", "protected area"]
abstract: "

1. A central aim in conservation is to maximise the long-term persistence of biodiversity. To fulfill this aim, reserve networks are used to safeguard biodiversity patterns (eg. species, populations) and processes (eg. evolutionary processes). Reserve selection is often formulated as an optimisation problem to identify cost-effective solutions. These formulations often use \"amount targets\" to specify the minimum amount habitat that must be preserved for each species. Although such targets are well suited for securing biodiversity patterns, they are not well suited for obtaining a representative sample of the biodiversity processes that play out across species' ranges. Currently, decision makers lack the tools to develop prioritisations that preserve both biodiversity patterns and processes.

2. To address this issue, we developed the \\texttt{rapr R} package. This \\texttt{R} package provides a toolkit to guide reserve selection using explicit targets for biodiversity processes. Using a novel formulation of the reserve selection problem, users can set \"space targets\" to secure a representative sample of each species within an attribute space to preserve biodiversity processes (eg. an attribute space based on genetic data to preserve evolutionary processes). We explored the functionality of this \\texttt{R} package using a simulation study and two case studies. In each study, we generated prioritisations using amount targets--to represent conventional reserve selection methods--and compared them to prioritisations generated using both amount and space targets.

3. We demonstrated how explicitly considering biodiversity processes in reserve selection can affect the resulting solution. The results from the simulation study suggest that most species could benefit from the explicit use of space-based targets--especially species with multi-model distributions in an attribute space. The results from both first case studies demonstrated that explicitly considering space-based targets for biodiversity processes can result in more effective prioritisations.

4. The \\texttt{rapr R} package provides a unified framework for biodiversity patterns and species-specific processes in conservation planning. Overall, we found strong evidence that prioritisations developed  using amount targets may fail to secure  that can provide better outcomes than just considering biodiversity patterns.
"
output:
  rmarkdown::pdf_document:
    toc: false
    keep_tex: yes
    fig_caption: yes
    includes:
      in_header: preamble-latex.tex
fontsize: 11pt
documentclass: article
bibliography: references.bib
csl: reference-style.csl
vignette: >
  %\VignetteIndexEntry{rapr}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

@@

## preliminary processing
for (i in seq_len(nlayers(cs1.bioclim.RAST)))
	cs1.bioclim.RAST@layers[[i]]@file@name  <- paste0(getwd(), strsplit(cs1.bioclim.RAST@layers[[i]]@file@name, 'rapr-manuscript')[[1]][[2]])

@@

<!--  title page -->
Short running title: Representative and adequate prioritisations \newline
Abstract word count: 322 / 350 \newline
Total word count: XXXXX / 8000 \newline
Number of references: 34 \newline
Number of figures: 8 \newline
Number of tables: 0 \newline
Number of textboxes: 0 \newline
Number of equations: 8 \newline
\newpage

# Introduction
<!-- conservation and reserve networks -->
The overarching aim of conservation is to maximise the long-term persistence of biodiversity [@r423; @r255]. To achieve this, conservation actions must preserve biodiversity patterns (eg. species, populations) and the processes that sustain them. One of the major tangible achievements of modern conservation has been the act of setting aside areas for preservation [@r440]. Reserve networks buffer species from threatening processes [eg. urbanisation; @r255] and set the stage for direct management interventions [eg. captive breeding and reintroduction programs; @r290]. However, the resources available for conservation action are limited, and so reserve networks must be sited in places that satisfy conservation objectives for minimum cost [@r255]. To achieve this, reserve selection is often formulated as an optimisation problem and then solved to identify cost-effective candidate reserve systems [prioritisations; @r255].

<!-- biodiversity processes are important -->
To fulfill the overarching aims of conservation, reserve networks must preserve both ecological and evolutionary processes [@r3; @r255]. Ecological processes, such as predator-prey interactions, pollination, and decomposition, are required for biodiversity to persist over short time-scales. Typically, they operate over small geographic domains--with exceptions such as migration and refugial habitats--and can be preserved using suitably large planning units [@r446] that each contain a discrete unit of habitat [@r448]. On the other hand, evolutionary processes are required for biodiversity to persist over long time-scales, and they typically operate over large geographic domains. Protected areas must preserve adaptive evolutionary processes to foster resilience against environmental change [eg. climate change; @r247]. Protected areas must also preserve neutral evolutionary processes [@r5; @r67; @r63]. These processes are associated with the breakdown of gene flow between populations. They are important for maintaining genetic diversity, and avoiding inbreeding depression. In recent decades, a wealth of data on biodiversity processes has become available to conservation planners [eg. bioclimatic, genetic, and trait data; @r33; @r453; @r460]. Yet this data is only rarely used to guide conservation planning [@r61]. This is due to that fact that existing reserve selection tools focus on preserving biodiversity patterns or processes--but not both.

<!-- parametrisation of biodiversity patterns and processes in conventional conservation planning tools -->
Many decision support tools have been developed to help identify cost-effective prioritisations [eg. \texttt{ ConsNet}, @r446; \texttt{ Marxan}, @r22; \texttt{ Zonation}, @r429]. Typically, these tools are used to deliver a prioritisation that preserves a set of target species (or features). Given a set of species, decision makers can use these tools to preserve biodiversity patterns by generating a prioritisation that secures an adequate proportion of each species' range. These tools, however, are not well-suited for biodiversity processes [but see @r203]. To preserve biodiversity processes, a prioritisation must preserve a representative sample of each species. For instance, to preserve predator-prey interactions, a prioritisation must preserve individuals from each predator and prey species in the same area. To preserve adaptive evolutionary processes, a prioritisation must preserve the adaptive landscape of each species--it must preserve individuals experiencing different selection pressures [@r63]. To preserve neutral evolutionary processes, a prioritisation must secure individuals descended from each of the genetic lineages that comprise each species [@r5]. To accommodate such information into a multi-species conservation planning exercise, species ranges must be partitioned into different groups [eg. using habitat data; @r2] as a pre-processing step to be compatible with existing tools. However, this approach assumes that the biodiversity processes that operate across a species' range can be split into discrete units. Yet data on biodiversity processes is often continuous and hyper-dimensional, and often cannot be reduced to a few categories without significant information loss [@r218].

<!-- aims of this paper -->
Today, one of the key issues in reserve selection is the lack of a unifying decision support tool that can accommodate data on biodiversity patterns and processes in a multi-species context. To fill this void, we present the \texttt{rapr R} package. This \texttt{R} package uses novel formulations of the reserve selection problem to provide decision makers with the tools to generate prioritisations that preserve biodiversity patterns and processes. We aim to define the concepts behind the problem formulations. Furthmore, we aim to explore the functionality of the \texttt{R} package by applying one of the formulations to a set of simulated species and two case studies.

# Methods
## PROBLEM FORMULATION
Biodiversity features are defined as the entity(s) that the prioritisation is required to preserve (eg. species, populations). Spatial attributes are defined as the variation across the species' range that the prioritisation is required to sample. They can be intrinsic (eg. genetic or trait variation) or extrinsic (eg. environmental variation) to the feature. These attributes should be related to the biodiversity processes that the decision maker aims to preserve.

Each attribute is conceptualised as a space. This space is termed an "attribute space". Each planning unit is thought to occupy a single point inside each space. For example, a decision maker may require a prioritisation that preserves populations along climatic gradients. To achieve this, the decision maker might use an "climatic" attribute space with dimensions relating to mean annual temperature ($^{\circ}$C) and precipitation (mm). Any given combination of temperature and precipitation may be conceived as a point in this environmental space. By associating planning units with climatic data, they can be mapped from geographic space to this environmental attribute space.

Demand points are points that also exist in an attribute space. They are designated by the decision maker to indicate regions of the attribute space that should be preserved in the prioritisation (see below for discussion on how demand points can be generated for real-world datasets). The amount of variation in the attribute space that a prioritisation secures is a function of the distance between each demand point and each selected planning unit in the attribute space. The shorter the distances between the demand points and the planning units; the better the prioritisation is at securing the variation in the spatial attribute. To convert these amounts to a proportion--a meaningful unit for a decision maker--the distances between the selected planing units and the demand points are scaled by the distances between the demand points to the centroid of the demand points. In any attribute space there may exist points that are impossible (eg. mean annual rainfall -5 mm), do not occur in the study area (eg. mean annual temperature 30$^{\circ}$C in Antarctica). Additionally, there may be some regions that are desirable for some features and undesirable for others (eg. conditions known to be outside the physiological tolerance of a species). Thus a different set of demand points and weights are used for each attribute space and each feature. By placing demand points in desirable regions of an attribute space for a given feature, the decision maker can ensure that prioritisations secure the feature in planning units with spatial attributes that are desirable for that feature.

To illustrate these concepts, we will briefly describe a conservation planning scenario example involving an attribute spaces and demand points. A decision maker may wish to develop a prioritisation for a single species. This species has four populations in the study area. However, the decision maker can only afford to preserve three population. The decision maker needs to preserve the adaptive landscape of the species, and preserve populations inhabiting different environmental conditions. To describe environmental variaiton, the decision maker obtained data on the environmental conditions (rainfall (ml) and temperature ($^{\circ}$C)) where each population was found. The decision maker then used this environmental data to construct a two-dimensional environmental attribute space. Next, the decision maker generated demand points as equi-distant points between the range of values where the populations were found. By comparing the distribution of the demand points to the distribution of the populations in the attribute space, the decision maker can identify a suitable prioritisation (Fig. 1). We can see that if the decision maker preserves both populations $A$ and $C$, they will effectively "double-up" on the same environmental characteristics, and in turn their waste resources. Instead, a more representative sample of the intra-specific variation could be preserved by securing populations $A$, $B$, and $D$. This example demonstrates how the inclusion of biodiversity processes can guide the reserve selection process.

The formulations used to express the reserve selection problem in the \texttt{rapr R} package are based on a combination of the \texttt{Marxan} reserve selection problem and the uncapacitated facility location problems [@r451]. Although the \texttt{rapr R} package provides two novel formulations, for brevity, we will define the simpler formulation--referred to as the unreliable formulation--below and define the more complex version--the reliable formulation--in the Supporting Information. The key difference between these two formulations is that the reliable formulation explicitly considers the probability that planning units are occupied when calculating the proportion of an attribute space sampled in a solution. All mathematical terms defined hereafter are described in Table S1. For convenience, the cardinality of sets will be denoted using the same symbol used to denote the variable.

Define $F$ to be the set of features (indexed by $f$). Let $J$ be a set of planning units (indexed by $j$). Also, let $A_j$ denote the area, and $C_j$ denote the cost of preserving planning unit $j \in J$. To assess the extent to which each feature is secured in a given prioritisation, let $q_{fj}$ denote the probability of feature $f$ occupying planning unit $j$. The level of fragmentation associated with a prioritisation is parametrised as the net exposed boundary length. Let the shared edges between each planning unit $j \in J$ and $k \in J$ be $e_{jk}$.

@@

## OPTIMISATION
The unreliable formulation is non-linear. However, the non-linear components can be linearised using existing techniques. The expression $X_j X_k$ in (eqn 3a) can be linearised using methods described by Beyer et al. [-@r426]. Linearised versions of the problems can be solved using commercial exact algorithm solvers. The \texttt{rapr} R package provides functions to express conservation planning data as optimisation problems using linearised versions of the unreliable and reliable formulations. These optimisation problems can then be solved to generate prioritisations using the commercial \texttt{Gurobi} software suite (\url{http://www.gurobi.com}). Presently, academics can obtain a license at no cost from the Gurobi website. After installing the \texttt{Gurobi} software suite, users will need to install the \texttt{Gurobi R} package.

@@

The uniform species was simulated with a constant probability of occupancy across the study area (Fig. 2a). The solution generated using `r sim.params.LST[[MODE]]$amount.target*100` % amount-based targets (Fig. 3a) selected `r filter(sim.spp.DF, species=='Uniform\nspecies', prioritisation=='Amount\ntargets')$n` planning units near the southern end of study area. This configuration is an artifact of the method used to solve this particular instance of the reserve selection problem. In fact, for this species, all prioritisations containing `r filter(sim.spp.DF, species=='Uniform\nspecies', prioritisation=='Amount\ntargets')$n` planning units are optimal when considering only `r sim.params.LST[[MODE]]$amount.target*100` % amount targets. In the absence of criteria to guarantee representativeness, reserve selection methods may or may not return solutions that secure a representative sample of the features. In this particular case, the solution does not secure a representative sample of the species' range (`r round(filter(sim.spp.DF, species=='Uniform\nspecies', prioritisation=='Amount\ntargets')$space.held,2)` % sampled).

The addition of a `r sim.params.LST[[MODE]]$space.target*100` % geographic target resulted in a solution that secured a representative sample of the uniform species' range (`r round(filter(sim.spp.DF, species=='Uniform\nspecies', prioritisation=='Amount &\nspace targets')$space.held,2)` % sampled; Fig. 3b). Since all planning units have cost the same amount and have an equal chance of being occupied, this solution has the same number of planning units as the solution generated using only amount targets (cf. Fig 3a). Although the use of amount and space based targets has addressed adequacy and representativeness objectives (respectively), they have resulted in a solution that is a highly fragmented solution.

@@

The bimodally distributed species was simulated to represent a species with a highly structured distribution, and it contains two isolated populations (Fig. 2c). The solution generated using just `r sim.params.LST[[MODE]]$amount.target*100` % amount targets assigned conservation effort to just one of the two populations (Fig. 3i). While this solution secured an adequate proportion of the species' distribution (`r  round(filter(sim.spp.DF, species=='Bimodal\nspecies', prioritisation=='Amount\ntargets')$amount.held,2)` % secured), it did not sample a representative proportion of the range (`r round(filter(sim.spp.DF, species=='Bimodal\nspecies', prioritisation=='Amount\ntargets')$space.held,2)` % sampled). The addition of boundary length modifiers to just amount targets resulted in a solution which sampled even less of the species distribution (`r round(filter(sim.spp.DF, species=='Bimodal\nspecies', prioritisation=='Amount target\nand BLM')$space.held,2)` % sampled; Fig. 3k). However, the addition of geographic space targets resolved these issues.

The solution generated using `r sim.params.LST[[MODE]]$amount.target*100` % amount and `r sim.params.LST[[MODE]]$space.target*100` % geographic targets preserved individuals in both populations (Fig. 3j). Although this solution required more planning units to fulfill both targets ($n=`r filter(sim.spp.DF, species=='Bimodal\nspecies', prioritisation=='Amount &\nspace targets')$n`$), this solution secured a representative sample of the bimodally distributed species (`r round(filter(sim.spp.DF, species=='Bimodal\nspecies', prioritisation=='Amount &\nspace targets')$space.held,2)` % range sampled). Similar to corresponding solutions generated for both the uniformly (Fig. 3b) and normally distributed species (Fig. 3f), this solution was also highly fragmented and we could obtain a more well-connected solution at the expense of selecting additional planning units ($n=`r filter(sim.spp.DF, species=='Bimodal\nspecies', prioritisation=='Amount & space\ntargets and BLM')$n`$; Fig. 3l). The results for the bimodally distributed species suggest that species with highly structured populations or strong variation between individuals could benefit the most from prioritisations that are generated using space-based targets.

## CASE STUDY 1
### Methods
Here we investigated how space-based targets can be used in a multi-species planning context to generate a prioritisation that sufficiently preserves the realised niche for several species. By preserving the populations in suitable habitats with different with environmental conditions, conservation planners can preserve the species' adaptive landscape and foster resilience against environmental change [@r63]. We selected Queensland, Australia as the study area. This region is ideal for exploring the potential of niche-based targets because it contains a broad range of different habitats. We obtained data for 19 bioclimatic variables across the region [at $30^{\prime \prime}$ resolution from \url{www.worldclim.org}; @r33] and subject them to a principal components analysis (using ArcMap 10.3.1). We used the first `r as.character(as.english(cs1.params.LST[[MODE]]$n.pc.layers))` principle components (cumulatively explained `r round(cs1.pca.DF[[4]][cs1.params.LST[[MODE]]$n.pc.layers],1)` % of the total variation) to characterise the environmental variation across the study area (Fig. 4).

We used four bird species in this case study: blue-winged kookaburra (_Dacelo leachii_), brown-backed honeyeater (_Ramsayornis modestus_), the brown falcon (_Falco berigora_), and the pale headed rosella (_Platycercus adscitus_). We used these bird species because they have different evolutionary histories and life-history strategies. We then mapped the extent of occurrence for each species (Fig. 5). To do this, we obtained observation data from the Atlas of Living Australia across the whole of Australia [using the \texttt{ALA4R R} package; @r453], spatially thinned the data to omit points within `r cs1.params.LST[[MODE]]$thin.distance/1000` km$^2$ of each other to ameliorate the effects of sampling bias [using the developmental version of the \texttt{spThin R} package; \url{www.github.com/mlammens/spThin}; @r454], and fit `r cs1.params.LST[[MODE]]$mcp.percent` % minimum convex polygons [using the \texttt{adehabitatHR R} package; @r455]. We used this method because it is reproducible using freely available data and moderately accurate.

@@

# Data accessibility
All data, code, and results are stored in an online repository (\url{www.github.com/paleo13/rapr-manuscript}) to replicate and valiate this study.

>>>>

---
title: "raptr: Representative and Adequate Prioritisation Toolkit in R"
author: |
  | Jeffrey O. Hanson$^1$, Jonathan R. Rhodes$^2$, Hugh P. Possingham$^1$, Richard A. Fuller$^1$
  | $^1$School of Biological Sciences, The University of Queensland, Brisbane, QLD, Australia
  | $^2$School of Geography, Planning and Environmental Management, The University of Queensland, Brisbane, QLD, Australia
  | Correspondance should be addressed to jeffrey.hanson@uqconnect.edu.au
date: "`r format(Sys.time(), '%d %B %Y')`"
tags: ["conservation", "biodiversity", "MILP", "reserve-selection", "optimisation", "protected area"]
abstract: "

1. An underlying aim in conservation planning is to maximise the long-term persistence of biodiversity. To fulfill this aim, the processes the sustain biodiversity must be conserved. Many biodiversity processes can be preserved by conserving a representative sample of each biodiversity feature (species or ecosystem). This idea has for decades been encapsulated in the principle of \"representativeness\". For example, representing the range of genetic diversity or ability to persist in a range of climatic conditions might be crucial for the long-term persistence of a species in a world where environmental change is accelerating. Yet current approaches overwhelmingly focus on \"adequacy\", the concept of ensuring a target proportion of the distribution of a biodiversity feature is reserved.

2. To address this issue, we developed the \\texttt{raptr: Representative and Adequate Prioritisation Toolkit in R}, a toolkit to guide reserve selection using explicit targets for representing biodiversity processes as well as area protection targets. Users set \"space targets\" to secure a representative sample of each feature within an attribute space that reflects a biodiversity process (eg. an attribute space expressing variation in genetic characteristics between individuals in an area, or variation in climatic conditions between inhabited areas). We explored the functionality of this package using a simulated study and two case studies. In each study, we generated solutions that secured a proportion of the species' distributions using amount targets--representing conventional reserve selection--and compared them with solutions generated using both amount and space targets.

3. We show that explicitly considering biodiversity processes in reserve selection can substantially change the solution. The simulations suggest that including representativeness in conservation planning is important where biodiversity features have multi-model distributions in an attribute space. The case studies show that explicitly considering space-based targets for biodiversity processes can return solutions that could be far more effective in achieving long term biodiversity persistence and do not require much more area than traditional approaches.

4. The \\texttt{raptr R} package provides a unified framework for achieving spatial conservation prioritisations that are both adequate and representative, resulting in a greater chance of achieving long-term biodiversity persistence than area-based planning alone.

"
output:
  rmarkdown::pdf_document:
    toc: false
    keep_tex: yes
    fig_caption: yes
    includes:
      in_header: preamble-latex.tex
fontsize: 11pt
documentclass: article
bibliography: references.bib
csl: reference-style.csl
---

@@

## preliminary processing
for (i in seq_len(nlayers(cs1.bioclim.RAST)))
	cs1.bioclim.RAST@layers[[i]]@file@name <- paste0(getwd(), strsplit(cs1.bioclim.RAST@layers[[i]]@file@name, 'raptr-manuscript')[[1]][[2]])

@@

<!--  title page -->
Short running title: Representative and adequate prioritisations \newline
Abstract word count: 350 / 350  
Total word count: XXXXX / 8000  
Number of references: 37
Number of figures: 8
Number of tables: 0  
Number of textboxes: 0  
Number of equations: 8  
\newpage

# Introduction
<!-- conservation and reserve networks -->
Perhaps the most fundamental aim of conservation is to maximise the long-term persistence of biodiversity at organisational levels [@r423; @r255]. To achieve this, conservation actions must preserve biodiversity patterns (eg. populations, species, ecosystems), but also crucially the processes that sustain them. One of the major tangible achievements of modern conservation has been the act of setting aside areas for preservation [@r440]. Reserve networks buffer species from gross threatening processes and set the stage for enhanced management interventions [@r465]. However, the resources available for conservation action are limited, and so reserve networks must be sited in places that satisfy conservation objectives for minimum cost [@r255]. To achieve this, reserve selection is often formulated as an optimisation problem and then solved to identify cost-effective candidate reserve systems [termed prioritisations; @r255]. Many decision support tools have been developed to help identify cost-effective solutions [eg. \texttt{ ConsNet}, @r446; \texttt{ Marxan}, @r22; \texttt{ Zonation}, @r429]. Typically, these tools are used to deliver a prioritisation that preserves a set of species of interest to the conservation planner (termed target species or features). Given a set of species, decision makers can use these tools to preserve biodiversity patterns by generating a prioritisation that secures an adequate proportion of each species' range.

<!-- biodiversity processes are important -->
However, to achieve effective conservation, reserve networks must do more than cover enough area. They must capture the processes that sustain biodiversity over the long term; they must capture both ecological and evolutionary processes [@r3; @r255]. Ecological processes, such as predator-prey interactions, pollination, and decomposition, are required for biodiversity to persist over short time-scales. Typically, they operate over small geographic domains--with exceptions such as migration and refugial habitats--and can be preserved using suitably large planning units [@r446] that each contain a discrete unit of habitat [@r448]. On the other hand, evolutionary processes are required for biodiversity to persist over long time-scales, and they typically operate over large geographic domains. Protected areas must represent adaptive evolutionary processes to foster resilience against environmental change [eg. climate change; @r247]. Protected areas must also preserve neutral evolutionary processes [@r5; @r67; @r63], arising from restriction of gene flow between populations. They are important for maintaining genetic diversity, and avoiding inbreeding depression. In recent decades, a wealth of data relevant to evolutionary processes has become available to conservation planners [eg. bioclimatic, genetic, and trait data; @r33; @r453; @r460]. Yet such data are rarely used to guide conservation planning [@r61]. This is perhaps in large part due to that fact that existing reserve selection tools focus on capturing either biodiversity patterns or processes--but not both.

<!-- parametrisation of biodiversity patterns and processes in conventional conservation planning tools -->
Existing conservation planning tools are generally not well-suited for representing biodiversity processes [but see @r203]. To preserve biodiversity processes, a prioritisation must capture a representative sample of each species. For instance, to preserve predator-prey interactions, a prioritisation must preserve individuals from each predator and prey species in the same area. To preserve adaptive evolutionary processes, a prioritisation must preserve the adaptive landscape of each species--populations experiencing different selection pressures [@r63]. To preserve neutral evolutionary processes, a prioritisation must secure individuals descended from each of the genetic lineages that comprise each species [@r5]. Previous attempts to accommodate such into multi-species conservation planning have partitioned species' ranges into different groups, for example at habitat discontinuities [e.g. @r2] as a pre-processing step to render a new set of "pseudo-species". However, this approach assumes that the biodiversity processes that operate across a species' range can be readily split into discrete units. Yet data on biodiversity processes is often continuous and hyper-dimensional, and often cannot be reduced to a few categories without significant information loss [@r218].

<!-- aims of this paper -->
One of the key issues in reserve selection is the lack of a unifying decision support tool that can accommodate data on biodiversity patterns and processes in a multi-species context. To begin to fill this gap, we present the \texttt{raptr R} package. This \texttt{R} package uses novel formulations of the reserve selection problem to provide decision makers with the tools to generate prioritisations that preserve both biodiversity patterns and processes. We aim to define the concepts behind the problem formulations. Furthermore, we aim to explore the functionality of the \texttt{raptr R} package by applying one of the formulations to a set of simulated species and two case studies.

# Methods
## PROBLEM FORMULATION
Biodiversity features are defined as the entities that the prioritisation is required to preserve (eg. species, ecosystems). Spatial attributes are defined as the variation across the species' range that the prioritisation is required to sample. They can be intrinsic (eg. genetic or trait variation) or extrinsic (eg. environmental variation) to the feature. These attributes should be related to the biodiversity processes that the decision maker aims to preserve.

Each attribute is conceptualised as a space. This space is termed an "attribute space". Each planning unit is thought to occupy a single point inside each space. For example, a decision maker may require a prioritisation that represents populations along climatic gradients. To achieve this, the decision maker might use an "climatic" attribute space with dimensions relating to mean annual temperature ($^{\circ}$C) and precipitation (mm). Any given combination of temperature and precipitation may be conceived as a point in this environmental space. By associating planning units with climatic data, they can be mapped from geographic space to this environmental attribute space.

Demand points are points that also exist in an attribute space. They are designated by the decision maker to indicate regions of the attribute space that the decision maker wishes to represent in the prioritisation (see below for discussion on how demand points can be generated for real-world datasets). The amount of variation in the attribute space that a prioritisation secures is a function of the distance between each demand point and each selected planning unit in the attribute space. The shorter the distances between the demand points and the planning units; the better the prioritisation is at securing the variation in the spatial attribute. To convert these amounts to a proportion--a meaningful unit for a decision maker--the distances between the selected planing units and the demand points are scaled by the distances between the demand points to the centroid of the demand points. In any attribute space there may exist points that are impossible (eg. mean annual rainfall -5 mm), or do not occur in the study area (eg. mean annual temperature 30$^{\circ}$C in Antarctica). Additionally, there may be some regions that are desirable for some features and undesirable for others (eg. conditions known to be outside the physiological tolerance of certain species). Thus a different set of demand points and weights are used for each attribute space and each feature. By placing demand points in desirable regions of an attribute space for a given feature, the decision maker can ensure that prioritisations secure the feature in planning units with spatial attributes that are desirable for that feature.

To illustrate these concepts, consider an example conservation planning scenario example involving an attribute space and demand points. We wish to develop a prioritisation for a single species that has four populations in the study area. However, we can only afford to preserve three populations. We aim preserve the adaptive landscape of the species, and to achieve this, we will preserve populations inhabiting different environmental conditions. To describe environmental variation, we obtain data on the environmental conditions (rainfall (mm) and temperature ($^{\circ}$C)) where each population is found. These environmental data are then used to construct a two-dimensional environmental attribute space. Next, we generated demand points as equi-distant points between the range of values where the populations were found. By comparing the distribution of the demand points to the distribution of the populations in the attribute space, we can identify a suitable prioritisation (Fig. 1). We can see that preserving both populations $A$ and $C$, effectively "doubles-up" on the same environmental characteristics, resulting in considerable redundancy in the solution. Instead, a more representative sample of the intra-specific variation could be captured by securing populations $A$, $B$, and $D$. This example demonstrates how the inclusion of biodiversity processes can guide the reserve selection process.

The formulations used to express the reserve selection problem in the \texttt{raptr R} package are based on a combination of the \texttt{Marxan} reserve selection problem and the uncapacitated facility location problems [@r451]. Although the \texttt{raptr R} package provides two novel formulations, for brevity, we will define the simpler formulation--referred to as the unreliable formulation--below and define the more complex version--the reliable formulation--in the Supporting Information S1. These formulations are named after the unreliable and reliable facility location formulations from which they are based upon [@r16]. The key difference between these two formulations is that the reliable formulation explicitly considers the probability that planning units are occupied when calculating the proportion of an attribute space sampled in a solution. All mathematical terms defined hereafter are described in Table S1. For convenience, the cardinality of sets will be denoted using the same symbol used to denote the variable.

Define $F$ to be the set of features one wishes to conserve (indexed by $f$). Let $J$ be a set of planning units (indexed by $j$). Also, let $A_j$ denote the area, and $C_j$ denote the cost of preserving planning unit $j \in J$. To assess the extent to which each feature is secured in a given prioritisation, let $q_{fj}$ denote the probability of feature $f$ occupying planning unit $j$. The level of fragmentation associated with a prioritisation is parametrised as the net exposed boundary length. Let the shared edges between each planning unit $j \in J$ and $k \in J$ be $e_{jk}$.

@@

## OPTIMISATION
The unreliable formulation is non-linear. However, the non-linear components can be linearised using existing techniques. The expression $X_j X_k$ in (eqn 3a) can be linearised using methods described by Beyer et al. [-@r426]. Linearised versions of the problems can be solved using commercial exact algorithm solvers. The \texttt{raptr} R package provides functions to express conservation planning data as optimisation problems using linearised versions of the unreliable and reliable formulations. These optimisation problems can then be solved to generate prioritisations using the commercial \texttt{Gurobi} software suite (\url{http://www.gurobi.com}). Presently, academics can obtain a license at no cost from the Gurobi website. After installing the \texttt{Gurobi} software suite, users will need to install the \texttt{Gurobi R} package.

@@

The uniform species was simulated to occur with a constant probability of occupancy across the study area (Fig. 2a). The solution generated using `r sim.params.LST[[MODE]]$amount.target*100` % amount-based targets (Fig. 3a) selected `r filter(sim.spp.DF, species=='Uniform\nspecies', prioritisation=='Amount\ntargets')$n` planning units near the southern end of study area. This configuration is an artifact of the method used to solve this particular instance of the reserve selection problem. In fact, for this species, all prioritisations containing `r filter(sim.spp.DF, species=='Uniform\nspecies', prioritisation=='Amount\ntargets')$n` planning units are optimal when considering only `r sim.params.LST[[MODE]]$amount.target*100` % amount targets. In the absence of criteria to guarantee representativeness, reserve selection methods may or may not return solutions that secure a representative sample of the features. In this particular case, the solution does not secure a representative sample of the species' range (`r round(filter(sim.spp.DF, species=='Uniform\nspecies', prioritisation=='Amount\ntargets')$space.held,2)` % sampled).

The addition of a `r sim.params.LST[[MODE]]$space.target*100` % geographic target resulted in a solution that secured a representative sample of the uniform species' range (`r round(filter(sim.spp.DF, species=='Uniform\nspecies', prioritisation=='Amount &\nspace targets')$space.held,2)` % sampled; Fig. 3b). Since all planning units have equal cost and an equal chance of being occupied, this solution has the same number of planning units as the solution generated using only amount targets (cf. Fig 3a). Although the use of amount and space based targets has addressed adequacy and representativeness objectives (respectively), they have resulted in a highly fragmented solution.

@@

The bimodally distributed species was simulated to represent a species with a highly structured distribution, and it contains two distinct populations (Fig. 2c). The solution generated using just `r sim.params.LST[[MODE]]$amount.target*100` % amount targets assigned conservation effort to just one of the two populations (Fig. 3i). While this solution secured an adequate proportion of the species' distribution (`r round(filter(sim.spp.DF, species=='Bimodal\nspecies', prioritisation=='Amount\ntargets')$amount.held,2)` % secured), it did not sample a representative proportion of the range (`r round(filter(sim.spp.DF, species=='Bimodal\nspecies', prioritisation=='Amount\ntargets')$space.held,2)` % sampled). The addition of boundary length modifiers to just amount targets resulted in a solution which sampled even less of the species distribution (`r round(filter(sim.spp.DF, species=='Bimodal\nspecies', prioritisation=='Amount target\nand BLM')$space.held,2)` % sampled; Fig. 3k). However, the addition of geographic space targets resolved these issues.

The solution generated using `r sim.params.LST[[MODE]]$amount.target*100` % amount and `r sim.params.LST[[MODE]]$space.target*100` % geographic targets included planning units from both populations (Fig. 3j). Although this solution required more planning units to fulfill both targets ($n=`r filter(sim.spp.DF, species=='Bimodal\nspecies', prioritisation=='Amount &\nspace targets')$n`$), this solution secured a representative sample of the bimodally distributed species (`r round(filter(sim.spp.DF, species=='Bimodal\nspecies', prioritisation=='Amount &\nspace targets')$space.held,2)` % range sampled). Similar to corresponding solutions generated for both the uniformly (Fig. 3b) and normally distributed species (Fig. 3f), this solution was also highly fragmented and we could obtain a more well-connected solution at the expense of selecting additional planning units ($n=`r filter(sim.spp.DF, species=='Bimodal\nspecies', prioritisation=='Amount & space\ntargets and BLM')$n`$; Fig. 3l). The results for the bimodally distributed species suggest that species with highly structured populations or strong variation between individuals could benefit the most from prioritisations that are generated using space-based targets.

## CASE STUDY 1
### Methods
We investigated how space-based targets can be used in a multi-species planning context to generate a prioritisation that sufficiently preserves the realised niche for several species. By preserving the populations in suitable habitats with different with environmental conditions, conservation planners can preserve the species' adaptive landscape and foster resilience against environmental change [@r63]. We selected Queensland, Australia as the study area. This region is ideal for exploring the potential of niche-based targets because it contains a broad range of different habitats. We obtained data for 19 bioclimatic variables across the region [at $30^{\prime \prime}$ resolution from \url{www.worldclim.org}; @r33] and subjected them to a principal components analysis (using ArcMap 10.3.1). We used the first `r as.character(as.english(cs1.params.LST[[MODE]]$n.pc.layers))` principal components (cumulatively explained `r round(cs1.pca.DF[[4]][cs1.params.LST[[MODE]]$n.pc.layers],1)` % of the total variation) to characterise the environmental variation across the study area (Fig. 4).

We selected four bird species that span a range of different evolutionary histories, distributions and ecologies: blue-winged kookaburra (_Dacelo leachii_), brown-backed honeyeater (_Ramsayornis modestus_), brown falcon (_Falco berigora_), and pale headed rosella (_Platycercus adscitus_). We then mapped the extent of occurrence for each species (Fig. 5). To do this, we obtained observation data from the Atlas of Living Australia across the whole of Australia [using the \texttt{ALA4R R} package; @r453], spatially thinned the data to omit points within `r cs1.params.LST[[MODE]]$thin.distance/1000` km$^2$ of each other to ameliorate the effects of sampling bias [using the developmental version of the \texttt{spThin R} package; \url{www.github.com/mlammens/spThin}; @r454], and fit `r cs1.params.LST[[MODE]]$mcp.percent` % minimum convex polygons [using the \texttt{adehabitatHR R} package; @r455]. We used this method because it is trivially reproducible using freely available data.

@@

# Data accessibility
All data, code, and results are stored in an online repository (\url{www.github.com/paleo13/raptr-manuscript}) to permit replication and validation of this study.
